---
title: 'Simulating transmission scenarios'
teaching: 10 # teaching time in minutes
exercises: 5 # exercise time in minutes
---

:::::::::::::::::::::::::::::::::::::: questions 

- What model structure do I need?
- How do I read in a social contact matrix? 
- How do I run a model simulation?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

Using the R package `epidemics`, learn how to: 

- load an existing model structure,
- load an existing social contact matrix,
- run a model simulation using.

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

Introductory text

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

Inline instructor notes can help inform instructors of timing challenges
associated with the lessons. They appear in the "Instructor View"

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisties

List (and hyperlink) the lessons/packages which need to be covered before this lesson

:::::::::::::::::::::::::::::::::

## Model structures

Lesson content

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1 : can the learner run existing code

Load contact and population data from socialmixr::polymod

```r
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
contact_data
```

:::::::::::::::::::::::: solution 

## Output
 
```{r polymod_uk, echo = FALSE}
 polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
contact_data
```


:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::


## Contact matrix and demography

Lesson content

::::::::::::::::::::::::::::::::::::: callout
## Explainer
Add additional maths (or epi) content for novice learners

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 2 : edit code/answer a question

Load contact and population data for Poland from socialmixr::polymod using the following age bins:

+ [0,15)
+ [15, 50)
+ 50 +

:::::::::::::::::::::::: solution 

```{r polymod_poland}
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "Poland",
  age.limits = c(0, 15, 50),
  symmetric = TRUE
)
contact_data
```
:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Model parameters

Lesson content

## Running simulations

Lesson content


::::::::::::::::::::::::::::::::::::: keypoints 

- Summarise the key points of the lesson using bullet points


::::::::::::::::::::::::::::::::::::::::::::::::
