---
title: 'Choosing a model'
teaching: 10 # teaching time in minutes
exercises: 20 # exercise time in minutes

---

```{r setup, echo= FALSE}
library(epidemics);library(ggplot2)
```


:::::::::::::::::::::::::::::::::::::: questions 

- How do I choose a model for my task?


::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn how to access the model library in `epidemics`
- Understand the model requirements for a question 

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites
+ Complete tutorial 'Simulating transmission'
:::::::::::::::::::::::::::::::::


## Introduction

Using mathematical models in outbreak analysis does not necessarily require developing a new model. There are existing models for different infections, interventions and transmission patterns which can be used to answer new questions. In this tutorial, we will learn how to choose an existing model to generate predictions for a given scenario.

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

The focus of this tutorial is understanding existing models to decide if they are appropriate for a defined question. 

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

### Choosing a model 

When deciding whether an existing model can be used, we must consider :

+ What is the infection/disease of interest? 
+ Is the modelling exercise for a disease with pandemic potential or a localised outbreak? 
+ What is the outcome of interest?
+ Will any interventions be modelled? 

A model may already exist for your study disease, or there may be a model for an infection that has the same transmission pathways and epidemiological features that can be used. 

Model structures differ for whether the disease has pandemic potential or not. We refer to such models as epidemic (pandemic potential) or outbreak (no pandemic potential). The choice between these two types of model will depend on the disease of interest and the question which will be answered with the modelling work.

The outcome of interest can be a feature of a mathematical model. It may be that you interested in the predicted numbers of infection through time, or in a specific outcome such as hospitalisations or cases of severe disease.

Finally, interventions such as vaccination may be of interest. A model may or may not have the capability to predict the impact of different interventions on different timescales (continuous time or at discrete time points). We will discuss interventions in detail in the next tutorial.


### Accessing the model library
 
The R package `epidemics` contains functions to run existing models. In `epidemics` we can choose from epidemic or outbreak models. Models are prefixed with either epidemic or outbreak, and suffixed by the infection name. 

*Outbreak* models focus on diseases that are not expected to have pandemic potential (such as Ebola). When predicted numbers of infection are small, stochastic variation in output can have an effect on whether an outbreak takes off or not. Outbreaks are usually smaller in magnitude than epidemics, so its often appropriate to use a stochastic model to characterise the uncertainty in the early stages of the outbreak.

*Epidemic* models are usually deterministic models focusing on diseases with pandemic potential. Epidemics are larger in magnitude than outbreaks and so a deterministic model is suitable as we have less interest in the stochastic variation in output. 

The names of available models can be found using the function `get_model_names()`.

```{r}
get_model_names(model_type = "epidemic")
```

Currently the only model type available is epidemic, further model types will be added in future. To find out more about each model read the documentation using `?epidemic_ebola` and the package [vignettes]{https://epiverse-trace.github.io/epidemics/articles}. Recall the check model equation questions we can consider when trying to understand an existing model. 


::::::::::::::::::::::::::::::::::::: checklist
### Check model equations

- How is transmission modelled? e.g. direct or indirect, airborne or vector-borne
- What interventions are modelled? 
- What state variables are there and how do they relate to assumptions about infection?

::::::::::::::::::::::::::::::::::::::::::::::::



## Challenge

::::::::::::::::::::::::::::::::::::: challenge

## What model?

You have been asked to explore the variation in numbers of infected individuals in the early stages of an Ebola outbreak. 

Which of the following models would be an appropriate choice for this task:

+  `epidemic_default`

+ `epidemic_ebola`

::::::::::::::::: hint

### HINT

Consider the following questions:

::::::::::::::::::::::::::::::::::::: checklist

+ What is the infection/disease of interest? 
+ Is the modelling exercise for a disease with pandemic potential or a localised outbreak?
+ What is the outcome of interest?
+ Will any interventions be modelled? 

::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::


::::::::::::::::: solution

### SOLUTION


+ What is the infection/disease of interest? **Ebola**
+ Is the modelling exercise for a disease with pandemic potential or a localised outbreak? **Localised outbreak**
+ What is the outcome of interest? **Number of infections**
+ Will any interventions be modelled? **No**

#### `epidemic_default`

A deterministic SEIR model with age specific direct transmission. The model is capable or predicting an Ebola type outbreak, but as the model is deterministic, we are not able to explore stochastic variation in the early stages of the outbreak.


#### `epidemic_ebola`

A stochastic SEIR model that was developed specifically for infection with Ebola. As this model is stochastic, it is the most appropriate choice to explore how variation in numbers of infected individuals in the early stages of an Ebola outbreak. 


:::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::::::::::::::



::::::::::::::::::::::::::::::::::::: keypoints 

- Existing models can be used for new questions
- Check that a model has appropriate assumptions about transmission, outbreak potential, outcomes and interventions 
::::::::::::::::::::::::::::::::::::::::::::::::

