---
title: 'Choosing a model'
teaching: 10 # teaching time in minutes
exercises: 20 # exercise time in minutes

---

```{r setup, echo= FALSE}
library(epidemics);library(ggplot2)
```


:::::::::::::::::::::::::::::::::::::: questions 

- How do I choose a model for my task?


::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn how to access the model library in `epidemics`
- Understand the model requirements for a question 

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites
+ Complete tutorial 'Simulating transmission'
:::::::::::::::::::::::::::::::::


## Introduction

Using mathematical models in outbreak analysis does not necessarily require developing a new model. There are existing models for different infections, interventions and transmission patterns which can be used to answer new questions. In this tutorial, we will learn how to choose an existing model to generate predictions for a given scenario.

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

Inline instructor notes can help inform instructors of timing challenges
associated with the lessons. They appear in the "Instructor View"

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

### Choosing a model 

When deciding whether an existing model can be used, we must consider :

+ What is the infection/disease of interest? 
+ Is the modelling exercise for a disease with pandemic potential or a localised outbreak?
+ What is the outcome of interest?
+ Will any interventions be modelled? 

A model may already exist for your study disease, or there may be a model for an infection that has the same transmission pathways and epidemiological features that can be used. 

Model structures differ for whether the disease has pandemic potential or not. We refer to such models as epidemic (pandemic potential) or outbreak (no pandemic potential). The choice between these two types of model will depend on the disease of interest and the question which will be answered with the modelling work.

The outcome of interest can be a feature of a mathematical model. It may be that you interested in the predicted numbers of infection through time, or in a specific outcome such as hospitalisations or cases of severe disease.

Finally, interventions such as vaccination may be of interest. A model may or may not have the capability to predict the impact of different interventions on different timescales (continuous time or at discrete time points). We will discuss interventions in detail in the next tutorial.


### Accessing the model library
 
The R package `epidemics` contains functions to run existing models. In `epidemics` we can choose from epidemic or outbreak models. Models are prefixed with either epidemic or outbreak, and suffixed by the infection name. The names of available models can be found using the function `get_model_names()`.

```{r}
get_model_names(model_type = "epidemic")
```

Currently the only model type available is epidemic, further model types will be added in future.

## Challenge

*Coming soon*

<!-- ::::::::::::::::::::::::::::::::::::: challenge -->

<!-- ## Generate predictions of XX caused by infection with XX -->

<!-- You have been asked to generate initial predictions of XX after an introduction of XX in a susceptible population. -->

<!-- You have been given the following information : -->

<!-- + the country of interest has a population of XX with contact matrix XX, -->
<!-- + the infection is introduced all age groups XX, with one in a million individuals infected -->

<!-- Select a model from the `epidemics` model library, specify the necessary inputs and create a visualisation of the predicted XX through time. -->

<!-- ::::::::::::::::: hint -->

<!-- ### HINT 1 : What model should I use? -->

<!-- You require a model which will generate predictions of infection without pandemic potential, so we need an *outbreak* model. -->

<!-- ::::::::::::::::: checklist -->

<!-- ### Check model equations -->

<!-- - How is transmission modelled? -->
<!-- - What interventions are modelled?  -->
<!-- - What state variables are there and how do they relate to assumptions about infection? -->

<!-- ::::::::::::::::::::::::::: -->


<!-- :::::::::::::::::::::: -->


<!-- ::::::::::::::::: hint -->

<!-- ### HINT 2 -->



<!-- :::::::::::::::::::::: -->

<!-- ::::::::::::::::: solution -->

<!-- ### SOLUTION  -->




<!-- :::::::::::::::::::::::::: -->

<!-- :::::::::::::::::::::::::::::::::::::::::::::::: -->



::::::::::::::::::::::::::::::::::::: keypoints 

- Existing models can be used for new questions
- Check that a model has appropriate assumptions about transmission, outbreak potential, outcomes and interventions 
::::::::::::::::::::::::::::::::::::::::::::::::

