---
title: "Week 1: Clean, validate linelist, and plot epicurves"
format: 
  html: # learners solutions
    embed-resources: true
  docx: default # learners practical
  gfm: default # instructors
keep-md: true
format-links: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {.content-hidden when-format="html"}

<!-- visible for instructors only -->
<!-- practical-week.md is generated from practical-week.qmd. Please edit that file -->
<!-- commit .md and .qmd files together -->

<!-- does not work for instructors text messages -->

:::

::: {.content-hidden when-format="docx"}

<!-- works for text on html and MD only -->

This practical is based in the following tutorial episodes:

- <https://epiverse-trace.github.io/tutorials-early/clean-data.html>
- <https://epiverse-trace.github.io/tutorials-early/validate.html>
- <https://epiverse-trace.github.io/tutorials-early/describe-cases.html>


:::


{{< include _welcome.qmd >}}


## Activity 1: Clean and standardize raw data

Get a clean and standardized data frame using the following available inputs:

- Raw messy data frame in CSV format

As a group, Write your answer to these questions:

- Diagnose the raw data. Write down all the cleaning tasks you plan to execute before writing the code.
- What time unit better describe the corresponding time span to calculate?
- Print the report: How would you communicate these results to a decision-maker?
- Compare: What differences you identify from other group outputs? (if available)

### Inputs

| Group | Data | Link | Calculate time span | Categorize time span |
|---|---|---|---|---|
| 1 | Small linelist | <https://epiverse-trace.github.io/tutorials-early/data/linelist-date_of_birth.csv> | Age until today | breaks = c(0, 20, 35, 60, 80) |
| 2 | Large linelist | <https://epiverse-trace.github.io/tutorials-early/data/covid_simulated_data.csv> | Delay from onset to death | breaks = c(0, 10, 15, 40) |
| 3 | Serology data | <https://epiverse-trace.github.io/tutorials-early/data/delta_full-messy.csv> | Time from last exposure | breaks = c(0, 30, 100, 600) |

::: {.content-visible when-format="docx"}

### Code chunk

```r
# Load packages ----------------------------------------------------------
library(cleanepi)
library(linelist)
library(incidence2)
library(tidyverse)


# Adapt the data dictionary ----------------------------------------------

# replace 'variable_name' when you have the information
dat_dictionary <- tibble::tribble(
  ~options,  ~values,            ~grp, ~orders,
       "1",   "male", "variable_name",      1L,
       "2", "female", "variable_name",      2L,
       "M",   "male", "variable_name",      3L,
       "F", "female", "variable_name",      4L,
       "m",   "male", "variable_name",      5L,
       "f", "female", "variable_name",      6L
)

dat_dictionary


# Read raw data ----------------------------------------------------------
dat_raw <- readr::read_csv(
  #<COMPLETE>
  )

dat_raw


# Clean and standardize data ---------------------------------------------

# how many cleanepi functions you used to get clean data?
dat_clean <- dat_raw %>%
  cleanepi::#<COMPLETE>

dat_clean


# Create time span variable ----------------------------------------------

# what time span unit better describe 'delay' from 'onset' to 'death'?
dat_timespan <- dat_clean %>%
  cleanepi::timespan(
    #<COMPLETE>
    #<COMPLETE>
    #<COMPLETE>
    span_column_name = "timespan_variable",
    span_remainder_unit = NULL
  ) %>%
  # skimr::skim(timespan_variable)
  # categorize the delay numerical variable
  dplyr::mutate(
    timespan_category = base::cut(
      x = timespan_variable,
      breaks = #<COMPLETE>, 
      include.lowest = TRUE,
      right = FALSE
    )
  )

dat_timespan


```

### Your answers

Group 1

| data | paste screenshot here |
|---|---|
| messy |  |
| clean |  |

Write your answer to the questions above:

```







```


------------------------

Group 2

| data | paste screenshot here |
|---|---|
| messy |  |
| clean |  |

Write your answer to the questions above:

```







```


------------------------

Group 3

| data | paste screenshot here |
|---|---|
| messy |  |
| clean |  |

Write your answer to the questions above:

```







```

:::




## Activity 2: Validate linelist and plot epicurve

Get a validated linelinst and incidence plot using the following available inputs:

- Clean data frame object

As a group, Write your answer to these questions:

- In the validation step, Do you need to allow for extra variables names and types?
- What is the most apprioriate time unit to aggregate the incidence plot? 
- Does using arguments like `fill`, `show_cases`, `angle`, `n_breaks` improves the incidence plot?
- Interpret: How would you communicate these results to a decision-maker?
- Compare: What differences you identify from other group outputs? (if available)

### Inputs

| Group | Data | Link | Calculate time span | Categorize time span |
|---|---|---|---|---|
| 1 | Small linelist | <https://epiverse-trace.github.io/tutorials-early/data/linelist-date_of_birth.csv> | Age until today | breaks = c(0, 20, 35, 60, 80) |
| 2 | Large linelist | <https://epiverse-trace.github.io/tutorials-early/data/covid_simulated_data.csv> | Delay from onset to death | breaks = c(0, 10, 15, 40) |
| 3 | Serology data | <https://epiverse-trace.github.io/tutorials-early/data/delta_full-messy.csv> | Time from last exposure | breaks = c(0, 30, 100, 600) |

::: {.content-visible when-format="docx"}

### Code chunk

```r
# Validate linelist ------------------------------------------------------

# activate Error message
linelist::lost_tags_action(action = "error")
# linelist::lost_tags_action(action = "warning")

# print tags types, names, and data to guide make_linelist
linelist::tags_types()
linelist::tags_names()
dat_timespan

# does the age variable pass the validation step?
dat_validate <- dat_timespan %>% 
  # tag variables
  linelist::make_linelist(
    #<COMPLETE>
    occupation = "timespan_category" # categorical var.
  ) %>% 
  # validate linelist
  linelist::#<COMPLETE> %>% 
  # test safeguard
  # dplyr::select(case_id, date_onset, sex)
  # INSTEAD
  linelist::tags_df()


# Create incidence -------------------------------------------------------

# what is the most appropriate time-aggregate (days, months) to plot?
dat_incidence <- dat_validate %>%  
  # transform from individual-level to time-aggregate
  incidence2::incidence(
    date_index = #<COMPLETE>,
    groups = "occupation", #OR any categorical variable
    interval = #<COMPLETE>,
    complete_dates = TRUE
  )


# Plot epicurve ----------------------------------------------------------

# does using arguments like 'fill', 'show_cases', 'angle', 'n_breaks' improves the plot?
dat_incidence %>% 
  plot(
    fill = "occupation", #<KEEP OR DROP>
    show_cases = TRUE, #<KEEP OR DROP>
    angle = 45, #<KEEP OR DROP>
    n_breaks = 5 #<KEEP OR DROP>
  )

# find plot() arguments at ?incidence2:::plot.incidence2()

```

### Your answers

Group 1

| data | paste screenshot here |
|---|---|
| validated |  |
| incidence plot |  |

Write your answer to the questions above:

```







```


------------------------

Group 2

| data | paste screenshot here |
|---|---|
| validated |  |
| incidence plot |  |

Write your answer to the questions above:

```







```


------------------------

Group 3

| data | paste screenshot here |
|---|---|
| validated |  |
| incidence plot |  |

Write your answer to the questions above:

```







```

:::




::: {.content-visible unless-format="docx"}

### Solution

<!-- visible for instructors and learners after practical (solutions) -->

#### Code

##### sample 1

```{r}
#| warning: false
#| eval: false

# Load packages ----------------------------------------------------------
library(cleanepi)
library(linelist)
library(incidence2)
library(tidyverse)


# Adapt the data dictionary ----------------------------------------------
# wait until have more information
dat_dictionary <- tibble::tribble(
  ~options,  ~values,        ~grp, ~orders,
       "1",   "male", "sex_fem_2",      1L,
       "2", "female", "sex_fem_2",      2L
)

dat_dictionary

# Read raw data ----------------------------------------------------------
dat_raw <- readr::read_csv(
  "https://epiverse-trace.github.io/tutorials-early/data/linelist-date_of_birth.csv"
)

dat_raw


# Clean and standardize data ---------------------------------------------

# what steps you required to have clean data?
dat_clean <- dat_raw %>%
  # standardize column names and dates
  cleanepi::standardize_column_names() %>%
  cleanepi::standardize_dates(
    target_columns = c(
      "date_of_admission",
      "date_of_birth",
      "date_first_pcr_positive_test"
    )
  ) %>%
  cleanepi::check_date_sequence(
    target_columns = c(
      "date_of_birth",
      "date_first_pcr_positive_test",
      "date_of_admission"
    )
  ) %>%
  # using data_dictionary requires valid missing entries
  cleanepi::replace_missing_values(
    target_columns = "sex_fem_2",
    na_strings = "-99"
  ) %>%
  cleanepi::clean_using_dictionary(dictionary = dat_dictionary) %>%
  cleanepi::remove_constants() %>%
  cleanepi::remove_duplicates(
    target_columns = c("study_id", "date_of_birth")
  )

dat_clean


# Create categorical variable --------------------------------------------

# what time span unit better describe age?
dat_timespan <- dat_clean %>%
  # calculate the age in 'years' and return the remainder in 'months'
  cleanepi::timespan(
    target_column = "date_of_birth",
    end_date = Sys.Date(),
    span_unit = "years",
    span_column_name = "age_in_years",
    span_remainder_unit = "months"
  ) %>%
  # skimr::skim(age_in_years)
  # categorize the age numerical variable
  dplyr::mutate(
    age_category = base::cut(
      x = age_in_years,
      breaks = c(0, 20, 35, 60, 80),
      include.lowest = TRUE,
      right = FALSE
    )
    # age_category = Hmisc::cut2(x = age_in_years,cuts = c(20,35,60))
  )

dat_timespan


# Validate linelist ------------------------------------------------------

# activate Error message
linelist::lost_tags_action(action = "error")
# linelist::lost_tags_action(action = "warning")

# print tags types, names, and data to guide make_linelist
linelist::tags_types()
linelist::tags_names()
dat_timespan

# does the age variable pass the validation step?
dat_validate <- dat_timespan %>%
  # tag variables
  linelist::make_linelist(
    id = "study_id",
    date_reporting = "date_first_pcr_positive_test",
    gender = "sex_fem_2",
    # age = "age_category", # does not pass validation
    age = "age_in_years",
    occupation = "age_category" # (downstream implications!)
  ) %>%
  # validate linelist
  linelist::validate_linelist() %>%
  # safeguard
  # TRY using
  # dplyr::select(study_id, sex_fem_2, age_category)
  # CONSEQUENCE
  # You get an ERROR notification due to loosing tags
  # INSTEAD
  # get a dataframe with all validated tags
  linelist::tags_df()

# relevant change: the variable names CHANGE to tag names!
# (can simplify downstream analysis!)
dat_validate


# Create incidence -------------------------------------------------------

# what is the most appropriate time-aggregate (days, months) to plot?
dat_incidence <- dat_validate %>%
  # transform from individual-level to time-aggregate
  incidence2::incidence(
    date_index = "date_reporting", #"date_first_pcr_positive_test",
    groups = "occupation", #"age_category", # change to sex, ...
    interval = "month", # change to days, weeks, ...
    complete_dates = TRUE # relevant to downstream analysis [time-series data]
  )

dat_incidence


# Plot epicurve ----------------------------------------------------------

# does using arguments like 'fill' or 'show_cases' improves the plot?
dat_incidence %>%
  plot(
    fill = "occupation", # "age_category",
    show_cases = TRUE,
    angle = 45,
    n_breaks = 5
  )

# find plot() arguments at ?incidence2:::plot.incidence2()
```

##### sample 2

```{r}
#| warning: false
#| eval: false

# Load packages ----------------------------------------------------------
library(cleanepi)
library(linelist)
library(incidence2)
library(tidyverse)


# Adapt the data dictionary ----------------------------------------------
# wait until have more information
dat_dictionary <- tibble::tribble(
  ~options,  ~values,  ~grp, ~orders,
       "1",   "male", "sex",      1L,
       "2", "female", "sex",      2L,
       "M",   "male", "sex",      3L,
       "F", "female", "sex",      4L,
       "m",   "male", "sex",      5L,
       "f", "female", "sex",      6L
)

dat_dictionary

# Read raw data ----------------------------------------------------------
dat_raw <- readr::read_csv(
  "https://epiverse-trace.github.io/tutorials-early/data/covid_simulated_data.csv"
)

dat_raw


# Clean and standardize data ---------------------------------------------

# what steps you required to have clean data?
dat_clean <- dat_raw %>%
  cleanepi::standardize_column_names() %>%
  cleanepi::standardize_dates(
    target_columns = c(
      "date_onset",
      "date_admission",
      "date_outcome",
      "date_first_contact",
      "date_last_contact"
    )
  ) %>%
  cleanepi::check_date_sequence(
    target_columns = c(
      "date_first_contact",
      "date_last_contact",
      "date_onset",
      "date_admission",
      "date_outcome"
    )
  ) %>%
  cleanepi::convert_to_numeric(target_columns = "age") %>%
  # dplyr::count(sex)
  # using data_dictionary requires valid missing entries
  cleanepi::replace_missing_values(
    target_columns = "sex",
    na_strings = "-99"
  ) %>%
  cleanepi::clean_using_dictionary(dictionary = dat_dictionary) %>%
  cleanepi::remove_constants() %>%
  cleanepi::remove_duplicates(
    target_columns = c("case_id", "case_name")
  )

dat_clean


# Create time span variable ----------------------------------------------

# what time span unit better describe delay from onset to death?
dat_timespan <- dat_clean %>%
  # calculate the time delay from 'onset' to 'death' in 'days'
  cleanepi::timespan(
    target_column = "date_onset",
    end_date = "date_outcome",
    span_unit = "days",
    span_column_name = "delay_onset_death",
    span_remainder_unit = NULL
  ) %>%
  # skimr::skim(delay_onset_death)
  # categorize the delay numerical variable
  dplyr::mutate(
    delay_category = base::cut(
      x = delay_onset_death,
      breaks = c(0, 10, 15, 40),
      include.lowest = TRUE,
      right = FALSE
    )
    # age_category = Hmisc::cut2(x = age_in_years,cuts = c(20,35,60))
  )

dat_timespan


# Validate linelist ------------------------------------------------------

# activate Error message
linelist::lost_tags_action(action = "error")
# linelist::lost_tags_action(action = "warning")

# print tags types, names, and data to guide make_linelist
linelist::tags_types()
linelist::tags_names()
dat_timespan

# does the age variable pass the validation step?
dat_validate <- dat_timespan %>%
  # tag variables
  linelist::make_linelist(
    id = "case_id",
    date_onset = "date_onset",
    gender = "sex",
    age = "age",
    outcome = "outcome",
    occupation = "delay_category" # (downstream implications!)
  ) %>%
  # validate linelist
  linelist::validate_linelist() %>%
  # safeguard
  # TRY using
  # dplyr::select(case_id, date_onset, sex)
  # CONSEQUENCE
  # You get an ERROR notification due to loosing tags
  # INSTEAD
  # get a dataframe with all validated tags
  linelist::tags_df()

# relevant change: the variable names CHANGE to tag names!
# (can simplify downstream analysis!)
dat_validate


# Create incidence -------------------------------------------------------

# what is the most appropriate time-aggregate (days, months) to plot?
dat_incidence <- dat_validate %>%
  # transform from individual-level to time-aggregate
  incidence2::incidence(
    date_index = "date_onset",
    groups = "outcome", #"age_category", # change to sex, ...
    interval = "day", # change to days, weeks, ...
    complete_dates = TRUE # relevant to downstream analysis [time-series data]
  )

dat_incidence


# Plot epicurve ----------------------------------------------------------

# does using arguments like 'fill' or 'show_cases' improves the plot?
dat_incidence %>%
  plot(
    angle = 45,
    n_breaks = 5
  )

# find plot() arguments at ?incidence2:::plot.incidence2()

```

##### sample 3

> Group 3 should investigate about how the argument `allow_extra = TRUE` us used in this howto entry <https://epiverse-trace.github.io/howto/analyses/describe_cases/cleanepi-linelist-incidence2-stratified.html>

```{r}
#| warning: false
#| eval: false

# Load packages ----------------------------------------------------------
library(cleanepi)
library(linelist)
library(incidence2)
library(tidyverse)


# Read raw data ----------------------------------------------------------
dat_raw <- readr::read_csv("https://epiverse-trace.github.io/tutorials-early/data/delta_full-messy.csv")

dat_raw


# Clean and standardize data ---------------------------------------------
dat_clean <- dat_raw %>%
  # cleanepi
  cleanepi::standardize_column_names() %>%
  cleanepi::standardize_dates(target_columns = "date") %>% #
  cleanepi::convert_to_numeric(target_columns = "exp_num") %>%
  cleanepi::check_date_sequence(
    target_columns = c("last_exp_date", "date")
  )

dat_clean


# Create time span variable ----------------------------------------------

# what time span unit better describe 'time' from 'last exposure' to 'date'?
dat_timespan <- dat_clean %>%
  # cleanepi::print_report()
  cleanepi::timespan(
    target_column = "last_exp_date",
    end_date = "date",
    span_unit = "days",
    span_column_name = "t_since_last_exp",
    span_remainder_unit = NULL
  ) %>%
  # skimr::skim(t_since_last_exp)
  # categorize the delay numerical variable
  dplyr::mutate(
    time_category = base::cut(
      x = t_since_last_exp,
      breaks = c(0, 30, 100, 600),
      include.lowest = TRUE,
      right = FALSE
    )
  )

dat_timespan


# Validate linelist ------------------------------------------------------

dat_validate <- dat_timespan %>%
  # tag with {linelist}
  linelist::make_linelist(
    id = "pid",
    occupation = "time_category",
    allow_extra = TRUE,
    last_exp_date = "last_exp_date",
    t_since_last_exp = "t_since_last_exp",
    last_vax_type = "last_vax_type"
  ) %>%
  # validate
  linelist::validate_linelist(
    allow_extra = TRUE,
    ref_types = linelist::tags_types(
      last_exp_date = c("Date"),
      t_since_last_exp = c("numeric"),
      last_vax_type = c("character"),
      allow_extra = TRUE
    )
  ) %>%
  # keep tags data frame
  linelist::tags_df()

dat_validate


# Create incidence -------------------------------------------------------

dat_incidence <- dat_validate %>%
    # aggregate
    incidence2::incidence(
      date_index = "last_exp_date",
      groups = "last_vax_type", 
      interval = "month", # change: "day" or "week" etc
      complete_dates = TRUE # relevant to downstream analysis [time-series data]
    )

dat_incidence


# Plot epicurve ----------------------------------------------------------

dat_incidence %>%
  # plot
  plot(
    fill = "last_vax_type" # change to groups variable
  )

# find plot() arguments at ?incidence2:::plot.incidence2()

```

#### Outputs

figure

table

#### Interpretation

Interpretation template:

+ ...

Interpretation Helpers:

- ...


# Continue your learning path

<!-- Suggest learners to Epiverse-TRACE documentation or external resources --->

Explore the downstream analysis you can do with {incidence2} outputs

- <https://www.reconverse.org/incidence2/doc/incidence2.html#sec:building-on-incidence2> 

You can use {epikinetics} to estimate Bayesian hierarchical modelling of antibody kinetics. Explore this sample code:

- <https://epiverse-trace.github.io/tutorials-early/epikinetics-statistics.html>

:::

# end